Notes on Validated Model Counting \\ Version of \today

### 2022-03-21 doc/mc-notes.tex
Added benchmarks
Generalize graphs

\documentclass{llncs}

\usepackage{latexsym}
\usepackage{times}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{marvosym}
%% Colored hyperlink 
\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
%% Colored hyperlink showing link in TT font
% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}

\bibliographystyle{splncs04}

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

@@

\newcommand{\interp}{\sigma}
\newcommand{\interpset}[1]{\Sigma_{#1}}
\newcommand{\mcount}{\mu}

@@

\newcommand{\subs}[2]{[#1/#2]}
\newcommand{\substrue}[1]{\subs{#1}{\tautology}}
\newcommand{\subsfalse}[1]{\subs{#1}{\false}}
\newcommand{\subsflip}[1]{\subs{#1}{\obar{#1}}}

\title{Notes on Validated Model Counting}

@@

@@

\date{\today}

@@

For interpretation $\interp$ and a Boolean formula $E$ over $X$, we
use the notation $\interp\subs{x}{E}$ to denote the intepretation
$\interp'$, such that $\interp'(y) = \interp(y)$ for all $y \not = x$
and $\interp'(x) = \interp(E)$.  In particular, the notation
$\interp\subsflip{x}$ indicates the interpretation in which the value
assigned to $x$ is complemented, while others remain unchanged.

@@

\begin{lemma}
\label{lemma:independent:split}
If Boolean formula $F$ is independent of variable $x$, then:
\begin{equation}
|\interpset{F}| \; = \; 2\cdot |\interpset{F\land x}| = \; 2\cdot |\interpset{F\land \obar{x}}| \label{eqn:independent}
\end{equation}  
\end{lemma}  
That is, the set of models of $F$ is split evenly between
those assigning $\tautology$ to $x$ and those assigning $\nil$ to $x$.

@@

We consider a single Boolean operation, known as
``If-Then-Else,'' or simply ``ITE.''  For Boolean values $a, b, c$,
the operation is defined as $\ite(a,b,c) = (a \land b) \lor (\obar{a}
\land c)$.  This single operation can be used to express several common Boolean operations:
\begin{eqnarray*}
\obar{a} & = & \ite(a, \nil, \tautology) \\
a \land b & = & \ite(a, b, \nil) \\
a \lor b & = & \ite(a, \tautology, b) \\
a \rightarrow b & = & \ite(a, b, \tautology) \\
\end{eqnarray*}

@@

\section{ITE Graphs}
Model counting is especially simple for a class of formulas we
call ``free ITE graphs.''  This representation can be obtained
directly from an Ordered Binary Decision Diagram (OBDD) representation
of a Boolean function~\cite{Bryant:1986}.

An {\em ITE graph} is defined to be a directed acyclic graph,
consisting of three node types:
\begin{description}
\item[Constant:] Corresponds to value $\tautology$ or $\nil$.  A constant node has no incoming arcs.
\item[Input:] Corresponds to one of the variables in $X$.   A variable node has no incoming arcs.
\item[Operator:] Represents an application of the $\ite$ operation.  An operator node has three incoming arcs, labeled \ifarg{}, \thenarg{}, and \elsearg{}, corresponding to the three arguments of the $\ite$ operation.
\end{description}

A {\em proper} ITE graph is one for which the incoming \ifarg{} arc for every operator node is
from an input node.  That is, the result of one $\ite$ operation can
serve as the ``then'' or ``else'' argument to another $\ite$
operation, but not for the ``if'' argument.

When describing proper ITE graphs, we refer to constant and input
nodes by their associated value or variable.  An operator node is described by an expression of the form
$\ite(x, v_t, v_e)$, where $v_t$ and $v_e$ are the
nodes corresponding to the \thenarg{} and \elsearg{} inputs, respectively.

We can define the function $\depend$ mapping each node
$v$ in a proper ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
\begin{eqnarray}
\depend(\tautology) & = & \emptyset \label{eqn:depend:true}\\
\depend(\nil) & = & \emptyset \label{eqn:depend:false}\\
\depend(x) & = & \{x \} \label{eqn:depend:variable}\\
\depend(\ite(x, v_t, v_e)) & = & \{ x \} \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
\end{eqnarray}

@@

A {\em free} ITE graph is a proper ITE graph satisfying the added
property that for every operator node $v = \ite(x, v_t, v_e)$, its
incoming \thenarg{} and \elsearg{} arcs must be from nodes such that
$x \not \in \depend(v_t)$ and $x \not \in \depend(v_e)$.

Free ITE graphs correspond directly to a class of binary decision diagrams known as {\em free} BDDs (FBDDs)~\cite{gergov:ieeetc:1994}.  Ordered BDDs are a subclass of FBDDs.

\section{Model Counting Free ITE Graphs}

While model counting is a difficult problem for arbitrary
formulas, it readily be computed when $F$ takes the form of a free
ITE graph.

\begin{theorem}
\label{theorem:count}
Letting $n = |X|$, the following recursive formula holds for any node in a free ITE graph
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount(\ite(x, v_t, v_e)) & = & \frac{\mcount(v_t) + \mcount(v_e)}{2} \label{eqn:count:ite}
\end{eqnarray}  
\end{theorem}
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemma~\ref{lemma:independent:split}, (\ref{eqn:ite:interp}),
Lemma~\ref{lemma:independent:dset}, and (\ref{eqn:depend:ite}).  That is, both
$v_t$ and $v_e$ are independent of $x$.  That implies that half of the
satisfying assignments $\interp$ for $v_t$ have have $\interp(x) =
\tautology$.  Similarly, half of the satisfying assignments $\interp$
for $v_e$ have $\interp(x) = \nil$.  Equation (\ref{eqn:count:ite}) then follows by (\ref{eqn:ite:interp}).

@@

>>>>

\documentclass{llncs}
\usepackage{latexsym}
\usepackage{times}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{marvosym}
%% Colored hyperlink 
\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
%% Colored hyperlink showing link in TT font
% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}

\bibliographystyle{splncs04}
%\bibliographystyle{abbrv}

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
%\newcommand{\tautology}{1}
%\newcommand{\nil}{0}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

@@

\newcommand{\interp}{\alpha}
\newcommand{\interpset}[1]{{\cal M}(#1)}
\newcommand{\mcount}{\mu}

@@

\newcommand{\subs}[2]{[#2/#1]}
\newcommand{\substrue}[1]{\subs{#1}{\tautology}}
\newcommand{\subsfalse}[1]{\subs{#1}{\false}}
\newcommand{\subsflip}[1]{\subs{#1}{\obar{#1}}}

\newcommand{\subspace}{{\cal S}}

\title{Notes on Validated Model Counting \\ Version of \today}

@@

\authorrunning{R. E. Bryant}

\titlerunning{Validated Model Counting}

@@

@@

For assignment $\interp$ and a Boolean formula $E$ over $X$, we
use the notation $\interp\subs{x}{E}$ to denote the assignment
$\interp'$, such that $\interp'(y) = \interp(y)$ for all $y \not = x$
and $\interp'(x) = \interp(E)$.  In particular, the notation
$\interp\subsflip{x}$ indicates the assignment in which the value
assigned to $x$ is complemented, while others remain unchanged.

@@

The following lemmas describe the primary abstractions used to perform
model counting without enumerating all solutions.

\begin{lemma}
  \label{lemma:negation}
  For any formula $F$:
  \begin{eqnarray*}
    \mcount(\neg F) & =  & 2^n - \mcount(F)
  \end{eqnarray*}
\end{lemma}

\begin{lemma}
  \label{lemma:product}
  For a partitioning of the variables $X$ into disjoint sets $X_1$ and $X_2$, and for formulas $F_1$ defined over $X_1$ and $F_2$ defined over $X_2$:
  \begin{eqnarray*}
    \mcount(F_1 \land F_2) &=& \mcount(F_1) \cdot \mcount(F_2)
  \end{eqnarray*}
\end{lemma}

\begin{lemma}
  \label{lemma:sum}
  For formulas $F_1$ and $F_2$ such that $\interpset{F_1} \cap \interpset{F_2} = \emptyset$:
  \begin{eqnarray*}
    \mcount(F_1 \lor F_2) &=& \mcount(F_1) + \mcount(F_2)
  \end{eqnarray*}
\end{lemma}

@@

We consider a single Boolean operation, known as
``If-Then-Else,'' or simply ``ITE.''  For Boolean values $a, b, c$,
the operation is defined as $\ite(a,b,c) = (a \land b) \lor (\neg a
\land c)$.  This single operation can be used to express several common Boolean operations:
\begin{eqnarray*}
\neg a & = & \ite(a, \nil, \tautology) \\
a \land b & = & \ite(a, b, \nil) \\
a \lor b & = & \ite(a, \tautology, b) \\
a \rightarrow b & = & \ite(a, b, \tautology) \\
\end{eqnarray*}

@@

\section{ITE Graphs}

An {\em ITE graph} describes a combinational logic circuit constructed from two-input multiplexors.
It is defined as a directed acyclic graph
with three node types:
\begin{description}
\item[Constant:] Corresponds to value $\tautology$ or $\nil$.  A constant node has no incoming arcs.
\item[Input:] Corresponds to one of the variables in $X$.   A variable node has no incoming arcs.
\item[Operator:] Represents an application of the $\ite$ operation.  An operator node has three incoming arcs, labeled \ifarg{}, \thenarg{}, and \elsearg{}, corresponding to the three arguments of the $\ite$ operation.
\end{description}

When describing ITE graphs, we refer to constant and input
nodes by their associated value or variable.  An operator node is described by an expression of the form
$\ite(v_i, v_t, v_e)$, where $v_i$, $v_t$, and $v_e$ are the
nodes corresponding to the \ifarg{}, \thenarg{}, and \elsearg{} inputs, respectively.

We can define the function $\depend$ mapping each node
$v$ in an ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
\begin{eqnarray}
\depend(\tautology) & = & \emptyset \label{eqn:depend:true}\\
\depend(\nil) & = & \emptyset \label{eqn:depend:false}\\
\depend(x) & = & \{x \} \label{eqn:depend:variable}\\
\depend(\ite(x, v_t, v_e)) & = & \depend(v_i) \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
\end{eqnarray}

@@

A {\em variable-partitioned} ITE (VPITE) graph is one satisfying the property
that for every operator node $v = \ite(v_i, v_t, v_e)$, the dependency
set of its \ifarg{} input most be disjoint from those of its
\thenarg{} and \elsearg{} inputs.  That is, both $\depend(v_i) \cap \depend(v_t) = \emptyset$ and
$\depend(v_i) \cap \depend(v_e) = \emptyset$.

\section{Model Counting VPITE Graphs}

While model counting is a difficult problem for arbitrary
formulas, it readily be computed when $F$ takes the form of a VPITE graph.

\begin{theorem}
\label{theorem:count}
Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2} \label{eqn:count:ite}
\end{eqnarray}  
\end{theorem}
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemmas~\ref{lemma:negation}--\ref{lemma:sum}, along with the definition of the ITE operator.

\section{Relation to Other Representations}

In general, there is a trade-off between the compactness of different
Boolean function representations versus the ease with which different
properties can readily be computed.  For example, Boolean formulas and
Boolean circuits can be very compact, but basic properties of these
representations, such as satisfiability, equivalence testing, and
model counting are NP-hard.  On the other hand all of these properties
can be computed in polynomial time for reduced, orderd binary decision
diagrams (ROBDDs)~\cite{Bryant:1986}, but they can grow exponentially
in relation to the representation of the function as a formula or
circuit.  ITE graphs lie somewhere between these two extremes.  Here
we examine how they relate to other representations and the implications
of these relations.

\subsection{Binary Decision Diagrams}

Binary Decision Diagrams~\cite{akers:ieeetc:1978} also represent Boolean formulas
as directed graphs consisting of ITE operators.  They can be seen as
equivalent to a class of graphs we will call {\em input-controlled} ITE
graphs.  These are ones for which the \ifarg{} input to every operator
node is a variable.

If we combine the properties of variable partitioned and input
control, we arrive at the representation known as {\em
  Free}~BDDs~\cite{gergov:ieeetc:1994}.  If we further restrict the
variables in a variable-controlled, input-partitioned ITE graph to
obey the same ordering along all paths formed in the graph, we arrive
at OBDDs.  An OBDD can be reduced by a simple set of rules to give an
ROBDD representation, and this form is canonical~\cite{Bryant:1986}.

\subsection{Sententential Decision Diagrams}

Darwiche uses the Sentential Decision Diagram (SDD)
representation~\cite{darwiche:ijcai:2011} as his framework for several
model counting programs.  An SDD can be seen as a combination of 1) a
generalization of VPITE graphs consisting of $k$-input multiplexors,
and 2) an ordering restriction on the variables.
That is, each operator node in an SDD is defined to
have $k$ data inputs and $k$ control inputs, where the Boolean
functions corresponding to the control inputs $p_1, p_2, \ldots, p_k$
satisfy $p_i \land p_j = \nil$ for $1 \leq i < j \leq k$ and
$\bigvee_{i=1,k} p_i = \tautology$.  The $k$ data inputs represent
Boolean functions $s_1, s_2, \ldots, s_k$.  In addition, the
dependency sets for the control and data inputs must be disjoint:
$\depend(p_i) \cap \depend(s_j) = \emptyset$ for $1 \leq i,j \leq k$.
Furthermore, the ordering of the variables in an SDD must obey a hierarchical decomposition
defined by
a tree structure known as a {\em vtree}.  Indeed, an SDD can be seen
as equivalent to a generalization of BDDs devised by McMillan, which
he termed {\em tree} BDDs~\cite{mcmillan:cav:1994}.

The graph part of an SDD can be readily translated into a VPITE graph
by expanding each $k$-input node into $k-1$ ITE nodes, giving an
output value defined as:
\begin{displaymath}
\ite(p_1, s_1, \ite(p_2, s_2, \ite( \cdots \ite(p_{k-1}, s_{k-1}, s_k)\cdots))
\end{displaymath}
If we express the size of an SDD as the number of branches, this
translation from SDD to ITE graph will expand the size by a constant
factor.

\section{Probabilistic Equivalence Checking}

One further feature of VPITE graphs is that there is a randomized
polynomial time (RP) algorithm for determining whether two graphs
represent equivalent Boolean functions.  This algorithm generalizes
the RP equivalence-testing algorithm for free BDDs described by Blum,
Chandra, and Wegman~\cite{blum:ipl:1980}.

@@

\section{A Model Counting Calculus}

Objective: Create a simple system for defining and reasoning about an
abstraction of Boolean formulas that suffices for tracking the number
of models.

Define a {\em counting subspace} $\subspace{} = \langle V, \alpha, k\rangle$, with:
\begin{itemize}
\item $V$ is a set of variables
\item $\alpha$ is a partial assignment to variables in $V$, represented as a set of literals.
\item $k$ is an integer, indicating the number of models for
  the formula within the specified subspace.  For $n = |V|$ and $m= |\var(\alpha)|$, it satisfies
  $0 \leq k \leq 2^{n-m}$.
\end{itemize}

\subsection{Subspace Primitives}
\begin{itemize}
\item
Universal subspace: $\langle \emptyset, \emptyset, 1 \rangle$.
\item
Null subspace: $\langle \emptyset, \emptyset, 0 \rangle$.
\item
  Positive literal: $\langle \{ x\}, \{x\}, 1\rangle$.
\item
  Negative literal: $\langle \{ x\}, \{\obar{x}\}, 1\rangle$.
\item
  Unused variable: $\langle \{ x\}, \emptyset, 2\rangle$.
\end{itemize}

\subsection{Basic Subspace Operations}

Assume $\subspace_1 = \langle V_1, \alpha_1, k_1\rangle$ and
$\subspace_2 = \langle V_2, \alpha_2, k_2\rangle$ are two subspaces

Subspace product $\subspace_1 \times \subspace_2$
\begin{itemize}
\item Condition: $V_1 \cap V_2 = \emptyset$
\item Result: $\langle V_1 \cup V_2, \alpha_1 \cup \alpha_2, k_1 \cdot k_2\rangle$.
\end{itemize}

Subspace Sum $\subspace_1 + \subspace_2$
\begin{itemize}
\item Conditions:
  \begin{itemize}
  \item $V_1 = V_2$.
  \item For some literal $l$ of variable $x \in V_1$ and assignment $\alpha$, the argument assignments satisfy
    $\alpha_1 = \alpha \cup \{l\}$ and $\alpha_2 = \alpha \cup \{ \obar{l} \}$.
  \end{itemize}
\item Result: $\langle V_1 - \{ x\}, \alpha, (k_1 + k_2)/2\rangle$.
\end{itemize}

\subsection{Derived Subspace Operations}

These useful operations can be constructed by combining primitive values and basic operations.

Expanding subspace $\subspace{} = \langle V, \alpha, k\rangle$:
\begin{itemize}
\item
  Condition: $x \not \in V$
\item
  Results:
  \begin{itemize}
    \item Positive expansion: $\langle V \cup \{ x \}, \alpha \cup \{ x \}, k \rangle$.
    \item Negative expansion: $\langle V \cup \{ x \}, \alpha \cup \{ \obar{x} \}, k \rangle$.
    \item Don't-care expansion: $\langle V \cup \{ x \}, \alpha , 2\cdot k \rangle$.
  \end{itemize}
\end{itemize}

A particularly useful expansion is to apply the don't-care expansion
with a set of variables $U$ such that $U \cap V = \emptyset$.  This
will give the subspace $\langle U \cup V, \alpha, 2^{|U|} \cdot
k\rangle$.
